---
title: "BIOS 545 Graphics Part 1"
author: Steve Pittard, <wsp@emory.edu>
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Intro

- R has a powerful environment for visualization of scientific data
- It provides publication quality graphics, which are fully programmable
- Easily reproducible
- Full LaTeX and Sweave support
- Lots of packages and functions with built-in graphics support 
- On-screen graphics
- Can save graphs to Postscript, PDF, jpeg, png, SVG

### History

- R Graph Gallery http://gallery.r-enthusiasts.com/

- R Graphic Manual and Gallery http://rgm2.lab.nig.ac.jp/RGM2/images.php?show=all&pageID=2087

- Grid Graphics - l http://www.stat.auckland.ac.nz/~paul/RGraphics/rgraphics.html

- Lattice Graphics - 
http://lmdvr.r-forge.r-project.org/figures/figures.html 

- ggplot2 and the tidyverse - https://ggplot2.tidyverse.org/

http://faculty.ucr.edu/~tgirke/HTML_Presentations/Manuals/Rgraphics/Rgraphics.pdf

### Levels

R graphics can be confusing because there are no less than 4 different systems. Let's list them out here and talk about which one(s) to use.

##### Low-Level Capability
  - Base Graphics (Has Low and High Level functions) 
  - Grid Graphics
  
##### High-Level Capability
  - Lattice Graphics 
  - ggplot2

### Base Graphics

- Oldest and most commonly used
- Uses a "pen-on-paper" model. You can only draw on top of the object. Cannot erase, modify, or delete what has already been drawn.
- Has both high and low level plotting routines (unique to Base) 
- Base graphics are fast.
- Lots of documentation and "google" support

### Grid Graphics

- Developed in 2000 by Paul Murrell
- Provides a rich set of graphics primitives
- Uses a system of objects and view ports to make complex objects easier.
- You will almost never use this directly unless you want to do in- depth programming

### Lattice package

- Developed by Deepayan Sarkar to implement the trellis graphics system described in "Visualizing Data" by Cleveland.

- Easy to create conditioned plots with automatic creation of axes, legends, and other annotations

- Usually considered to be an improvement over Base graphics as it relates to panels

```{r lattice1}
library(lattice)
xyplot(mpg~wt | factor(am,labels=c("Auto","Manual")), data=mtcars)
```

### ggplot2

- Developed starting in 2005 by Hadley Wickham

- ggplot2 is an implementation of Leland Wilkinson's Grammar of Graphics--a general scheme for data visualization which breaks up graph into semantic components such as scales and layers.

- ggplot2 can serve as a replacement for the base graphics in R and contains a number of defaults for web and print display of common scales.

- Is said to be much slower than Base graphics but this isn't a major thing (in my opinion)

## Chart Types

* plot(x,y) where x and y are continuous:
  - X/Y, scatterplot, pairs, sunflower plots   
  
* plot(x,[y]) where x and y are categorical. Note that y can be optional
  - dotplot, barplot, stacked bar plot, pie chart   
  
* plot(x) where x is a single continuous variable:
  - dotplot, barplot, stripchart, boxplot, density, histogram, QQ Plot    

* plot(x,y) where one of x and y is continuous and the other is discrete
  - Side-by-Side dotplot and boxplot, notched boxplot   
  

### Base Graphics Foundations

```{r bg1}
plot(0:10, 0:10, type="n", xlab="X", ylab="Y", axes=FALSE)

abline(h=seq(0,10,2),lty=3,col="gray90") 

abline(v=seq(0,10,2),lty=3,col="gray90")

text(5,5, "Plot Stuff Here", col="red", cex=1.5) 

box("plot", col="red", lty = "dotted")

box("inner", col="blue", lty = "dashed") 

mtext("South Margin",1,cex=1.2,col="blue") 

mtext("West Margin",2,cex=1.2,col="green") 

mtext("North Margin",3,cex=1.2,col="orange") 

mtext("East Margin",4,cex=1.2,col="purple") 

title("An Example Plot")
```

Let's do some basic plotting. These two commands do the same thing. Given two vectors, x and y (of same length), do a scatterplot. Typically we would do something like "plot(x,y)". 

```{r bgmt1}
plot(mtcars$wt, mtcars$mpg, main="MPG vs. Weight")
```
We can also plot a single variable:

```{r}
plot(mtcars$mpg, main="MPG", type="l", xlab="Car Number",ylab="MPG")
plot(mtcars$mpg, main="MPG", type="b", xlab="Car Number",ylab="MPG")
```

 We can also plot a single variable:
 
```{r}
boxplot(mtcars$mpg, main="BoxPlot of MPG", ylab="MPG",col="blue",notch=TRUE)
boxplot(mtcars$mpg,main="BoxPlot of MPG",ylab="MPG",col="red")
```
 
 How can I get two plots to be on the same page ?
 
```{r}
par(mfrow=c(1,2)) # One row and two columns
plot(mtcars$mpg, main="MPG", type="l", xlab="Car Number",ylab="MPG")
plot(mtcars$mpg, main="MPG", type="b", xlab="Car Number",ylab="MPG")
par(mfrow=c(1,1))

```
 We can create panels:
 
 
```{r}
par(mfrow=c(2,2)) 

plot(mtcars$mpg,main="MPG",xlab="Car",ylab="MPG",type="p") 

plot(mtcars$mpg,main="MPG",xlab="Car",ylab="MPG",type="l") 

plot(mtcars$mpg,main="MPG",xlab="Car",ylab="MPG",type="h") 

plot(mtcars$mpg,main="MPG",xlab="Car",ylab="MPG",type="o") 

legend("topleft",legend=c("Test Legend"),cex=0.8)

par(mfrow=c(1,1))
```
 
 We usually take this approach when we want to plot data across different categories. Like the mpg vs weight across cylinder types. We have three unique cylinder values:

```{r}
unique(mtcars$cyl)   # We have three categories so create 3 plots

par(mfrow=c(1,3)) # One row and three columns

fourcyl <- mtcars[mtcars$cyl == 4,] 

sixcyl <- mtcars[mtcars$cyl == 6,] 

eightcyl <- mtcars[mtcars$cyl == 8,]

plot(fourcyl$wt, fourcyl$mpg, main = "MPG vs Wt 4 Cyl", ylim=c(0,40))

plot(sixcyl$wt, sixcyl$mpg, main = "MPG vs Wt 6 Cyl", ylim=c(0,40))

plot(eightcyl$wt, eightcyl$mpg, main = "MPG vs Wt 8 Cyl", ylim=c(0,40))

par(mfrow=c(1,1)) # Reset the plot window
```

We could automate this using the "split" approach:

```{r}
par(mfrow=c(1,3)) # One row and three columns 

mysplits <- split(mtcars, mtcars$cyl)
for (ii in 1:length(mysplits)) {
  plot(mysplits[[ii]]$wt, mysplits[[ii]]$mpg, 
       ylim = c(0,40),main=paste("MPG vs weight for",names(mysplits[ii])))
}

# Better yet we could make this into a function
cyl.plot <- function(df, fac, numrows=1, numcols=3) { 
  par(mfrow=c(numrows,numcols))
  mysplits <- split(df,fac)
  for (ii in 1:length(mysplits)) {
    plot(mysplits[[ii]]$wt, mysplits[[ii]]$mpg, 
         ylim = c(0,40),
         main=paste("MPG vs weight for",names(mysplits[ii])))
  }
}

# Call the function   
cyl.plot(mtcars,mtcars$cyl)
```

We can set plot limits and add annotations

```{r}
plot(mtcars$wt, mtcars$mpg, 
     xlab = "Weight / 1,000", 
     ylab = "MPG", 
     xlim = c(0,10), 
     ylim = c(0,40))
```
```{r}
plot(mtcars$wt, mtcars$mpg, 
     xlab = "Weight / 1,000", 
     ylab = "MPG", 
     xlim = c(0,10), 
     ylim = c(0,40))

legend("topright", inset=0.05, "My Data", pch=1, col="black") 

# Could use specific coordinates also
legend(6.5,35, inset=0.05, "My Data", pch=1, col="black")


```

We could also put up our own grid using some "primitive" graphics functions:

```{r}
plot(mtcars$wt, mtcars$mpg,
     xlab = "Weight / 1,000",
     main = "MPG vs. WT") 

# Draws vertical dashed lines at 2,3,4,5 
abline(v=c(2,3,4,5),lty=2,col="gray")

# Horizontal lines at 10,15,20,25,30
abline(h=c(10,15,20,25,30), lty=2, col="gray") 

# Could do:
abline(v=2:5,lty=2,col="gray90") 
abline(h=seq(10,30,5),lty=2,col="gray")
```
We can change the plot character:

```{r}
plot(mtcars$wt, mtcars$mpg,
     main="MPG vs WT", col="red", 
     xlab="Weight in lbs/1,000",
     ylab="Miles per Gallon", 
     pch = 3)

legend("topright", inset=0.05, "My Data", pch = 3, col="red") 
grid(col="blue")
```
```{r}
example(pch)
```
### Case Study 1

We could also use information from a data frame to help us print different characters based on value. Like in mtcars. Let's plot MPG vs Weight but pick a different plot character based on Transmission Type. Here is one way to do it:

1. Create a blank plot that sets the limits and title
2. Extract records for automatic transmission into a data frame
3. Extract records for manual transmission into a data frame
4. Use the points command to plot these two different groups using a different
pch value

```{r}
# A null plot
plot(mtcars$wt, mtcars$mpg, type="n", main="MPG vs. Weight")

# Extract cars with an auto transmission (am = 0)
auto <- mtcars[mtcars$am == 0,] 

# Extract cars with a manual transmission (qm = 1)
manu <- mtcars[mtcars$am == 1,]

# Put up some points
points(auto$wt, auto$mpg, pch = 0) 
points(manu$wt, manu$mpg, pch = 1)

# Display a legend
legend("topright", inset=0.05, c("manual","auto"), pch = c(1,0))
```
But this would be working too hard. No programming is required. Just recognize that the plot characters are selected by a number from 0 to 25. We can exploit this:

```{r}
mtcars$am
```

We see that am is 0 or 1 which just so happen to also represent
valid print characters

```{r}
plot(mtcars$wt, mtcars$mpg, 
     pch=mtcars$am,
     main="MPG vs. Weight", 
     sub="Different plot chars")

legend("topright", inset=0.05, c("manual","auto"), pch = unique(mtcars$am))
```

```{r}
plot(mtcars$wt, mtcars$mpg, 
     pch=mtcars$am+2,
     main="MPG vs. Weight", 
     sub="Different plot chars")

legend("topright", inset=0.05, c("manual","auto"), pch = unique(mtcars$am)+2)
```
### Case Study 2

One of the strengths of Base Graphics is that we can build plots in layers. 

First, we initialize a "blank" plot using the plot command but we specify a 
type of "n". In this case "n" means "NULL". 

Let's then plot wt vs mpg and do it such that the records with a weight below the mean weight of the entire data set are in red and those above the mean weight are in blue.

```{r}
plot(mtcars$wt,mtcars$mpg,type="n", 
     xlab="Weight in lbs/1,000", ylab="MPG", main="MPG vs. Weight")
```

How is this useful ? Well we can add points or lines in stages. This allows us to plot things on an existing plot using specific colors or print characters.


```{r}
plot(mtcars$wt,mtcars$mpg,
     type="n",xlab="Weight in lbs/1,000", 
     ylab="MPG", main="MPG vs. Weight")

# Let's get records for each category
above.mean <- mtcars[mtcars$wt >= mean(mtcars$wt),] 
below.mean <- mtcars[mtcars$wt < mean(mtcars$wt),] 

# Use the points command to plot each group
points(below.mean$wt,below.mean$mpg,col="red")
points(above.mean$wt,above.mean$mpg,col="blue")

# Draw a vertical line where the mean(wt) is
abline(v=mean(mtcars$wt),lty=2,col="gray")
```
Unfortunately there is nothing in the existing data set that tells us if a given row's weight value is greater than or below the mean weight. We could handle this a couple of ways - one of which is to use our knowledge of for loops.

```{r}
colvec <- ifelse(mtcars$wt >= mean(mtcars$wt),"blue","red")
colvec
```

We could now use this information to our advantage:

```{r}
plot(mtcars$wt,mtcars$mpg,col=colvec)
```

We can add text to our plot with no problem.

```{r}
plot(mtcars$wt, mtcars$mpg, 
     main="Mileage vs. Car Weight", 
     xlab="Weight",
     ylab="Mileage", pch=18, col="blue")

text(mtcars$wt, mtcars$mpg, 
     row.names(mtcars),     # Get the row names
     cex=0.6,               # Scale the font size
     pos=4,                 # 1=below, 2=left, 3=above, 4=right
     col="red")
```

```{r}
plot(mtcars$wt, mtcars$mpg, 
     main="Mileage vs. Car Weight", 
     xlab="Weight",
     ylab="Mileage", pch=18, col="blue")

carlabs <- sapply(strsplit(row.names(mtcars)," "), function(x) x[[1]])

carlabs
```

```{r}
plot(mtcars$wt, mtcars$mpg, 
     main="Mileage vs. Car Weight", 
     xlab="Weight",
     ylab="Mileage", pch=18, col="blue")

carlabs <- sapply(strsplit(row.names(mtcars)," "), function(x) x[[1]])

text(mtcars$wt, mtcars$mpg, 
     carlabs,     # Get the row names
     cex=0.6,               # Scale the font size
     pos=4,                 # 1=below, 2=left, 3=above, 4=right
     col="red")
```
Let's look at a more involved annotation example. We'll use the same data:

```{r}
plot(mtcars$mpg ~ mtcars$wt,
     cex=0.8, pch=21,
     col="blue",bg="red", 
     xlab="Wt in Lbs/1,000", 
     ylab="Miles Per Gallon")

title(main="The mtcars data set wt vs. MPG")

# Next draw a vertical line at the mean of the weight
abline(v=mean(mtcars$wt),lty=2,col="blue")

# Next draw a horizontal line at the man of the MPG
abline(h=mean(mtcars$mpg),lty=2,col="blue")

# Draws a diamond at the common mean
points(mean(mtcars$wt), mean(mtcars$mpg),
       pch=23,col="black", bg="brown",cex=2)
```

```{r}
plot(mtcars$mpg ~ mtcars$wt,
     cex=0.8, pch=21,
     col="blue",bg="red", 
     xlab="Wt in Lbs/1,000", 
     ylab="Miles Per Gallon")

title(main="The mtcars data set wt vs. MPG")

# Next draw a vertical line at the mean of the weight
abline(v=mean(mtcars$wt),lty=2,col="blue")

# Next draw a horizontal line at the man of the MPG
abline(h=mean(mtcars$mpg),lty=2,col="blue")

# Draws a diamond at the common mean
points(mean(mtcars$wt), mean(mtcars$mpg),
       pch=23,col="black", bg="brown",cex=2)

# Let's put some custom text on the graph to indicate the mean.

text(mean(mtcars$wt),min(mtcars$mpg), 
     paste("Mean:",round(mean(mtcars$wt),2)),pos=4)

text(min(mtcars$wt),mean(mtcars$mpg),
     paste("Mean:",round(mean(mtcars$mpg),2)),adj=c(0,1))

# Note that this is basically equivalent to:

text(3.2,10.4, paste("Mean:",round(mean(mtcars$wt),2)),pos=4)

# text(2,20.09,paste("Mean:",round(mean(mtcars$mpg),2)))
```

### Custom Axes

Sometimes we want to draw an axis ourselves because R's defaults aren't what we want. Imagine a set of observations over time such as stock market activity. Here is a data frame you can read in that tracks actual stock market performance for Microsoft, (MSFT), for each trading day of the year 2014.

```{r}
url <- "https://raw.githubusercontent.com/steviep42/bios545_spring_2021/master/DATA.DIR/stock.data.14.csv"

msft <- read.csv(url,stringsAsFactors = FALSE) 
head(msft)
```

```{r}
plot(msft$High,type="l",main="High MSFT Price",
     xlab="Day of Year",ylab="Price in $") 
mtext("2014",3)
```

The day number is okay but we could also the actual dates as labels. But that could be a problem. First, we use the xaxt argument to suppress the printing of the x-axis

```{r}
plot(msft$High,type="l",main="High MSFT Price", 
     xlab="Day of Year",ylab="Price in $", xaxt="n")

axis(1,at=1:nrow(msft),labels=msft$Date)
```

That wasn't so good because the X-axis got really crowded. We can print labels for the x-axis every 30 days or so using this approach.

We could alter this to accommodate an arbitrary number of days and labels.

Notice how we generate sequence that we then use to index into the Dates.

```{r}
plot(msft$High,type="l",main="High MSFT Price", 
     xlab="Day of Year",ylab="Price in $", xaxt="n")

mtext("2014",3)

dseq <- seq(1,nrow(msft),30)
axislabs <- substr(msft$Date[dseq],6,10) 
axis(1, at=dseq, labels=axislabs, cex.axis=0.8)
```
### Working with Colors

```{r}
colors()
length(colors())
```

```{r}
grep("yellow",colors(),value=TRUE)

grep("purple",colors(),value=TRUE)
```

Get a copy of the PDF Color Chart from: http://research.stowers-institute.org/efg/R/Color/Chart/ColorChart.pdf

R also has some built in palettes that give you a color scheme from which to choose:

If we have some categories we want to look at we can easily visualize it. Barplots are for plotting tables. Let's count up all the cars by cylinder type from mtcars:

```{r}
table(mtcars$cyl)
```

Pass this to a barplot

```{r}
barplot(table(mtcars$cyl), axes=T, main = "Cylinder Barplot")
```

```{r}
barplot(table(mtcars$cyl), axes=T,
main = "Cylinder Barplot", col=heat.colors(3))
```

```{r}
(myt <- table(mtcars$cyl,mtcars$am))

```
```{r}
barplot(myt, legend = T, 
        beside = T, 
        col=heat.colors(3), main='Cylinder Count by Transmission Type')
```
```{r}
barplot(myt,
        legend = T, 
        beside = F, 
        col=heat.colors(3),
        main="Cylinder Count by TransmissionType", 
        ylim=c(0,25))
```

```{r}
barplot(myt,
        legend = T, 
        beside = F, 
        col=heat.colors(3),
        main="Cylinder Count by Transmission Type", 
        xlim=c(0,25),
        horiz=T)
```
Let's use our new found knowledge of colors

```{r}
mycols <- rainbow(2)
mycols
```

Remember that the transmission types are indicated by a 0 (auto) or 1 (manual). We need to take # this into account when indexing into the mycols vector.

```{r}
plot(mtcars$wt, mtcars$mpg, col = mycols[mtcars$am+1], pch=19)

legend("topright",c("Auto","Manual"),
       col = mycols,
       pch = 19)
```

Consumer graphs, like those found in newspapers or news magazines, have lots of "junk" attached to them, which, for a statistician, is unnecessary.
Here is an example found in a copy of the Atlanta Journal Constitution newspaper from some time last year. You can find these in many magazines and papers.

I wrote some R code using Base graphics to approximate this.This took a lot of work since the chart relies on intersecting lines, different colors, custom axes, arrows, and text annotations.

I don't enjoy doing things like this at all. A good plot should tell the story without all the extraneous annotations.

But for the public, this type of chart is standard. See the result on the next slide. It's not a perfect match. The x-axis labels need some more work but it's close enough. The code can be found at:

```{r}
# Reproduction of a consumer level graphic publication found in the 
# Atlanta Journal Constitution. 
# Steve Pittard wsp@emory.edu

# Tuition Data Set - Approximate values

Wyoming <- c(3242,3342,3442,3542,3642,3752,3880,3950,4100,4404)
Georgia <- c(3411,3630,4000,4200,4500,5000,6000,7600,7700,7823)
USAver  <- c(4962,5100,5800,6000,6200,6800,7300,8000,8200,8596)
NewHamp <- c(8085,8300,9000,9800,10150,10900,12000,13750,14500,14665)

tuition <- data.frame(rbind(Wyoming,Georgia,USAver,NewHamp))
ylims <- c(0,max(tuition[,10]))

plot(1:10,tuition[1,],type="n",lwd=2,col="green",
     ylab="",ylim=ylims,yaxt="n",bty="n",xaxt="n",
     xlab="Year",main="Tuition Growth",sub="Figures adjusted for 2014 dollars")
mcols <- c("#6E8B3D","blue","red","black")
polygon(c(1,1:10,10),c(0,tuition[4,],0),col="#EEE8AA")

points(1:10,tuition[1,],type="l",lwd=3.5,col="#6E8B3D")
points(1:10,tuition[2,],type="l",lwd=3.5,col="blue")
points(1:10,tuition[3,],type="l",lwd=3.5,col="red")
points(1:10,tuition[4,],type="l",lwd=3.5,col="black")

abline(v=1:10,col="white",lwd=1.5)
segments(1,seq(2500,15000,by=2500),10,seq(2500,15000,by=2500),lty=3,col="#6E8B3D")
xlabs <- c(paste(0,1:9,sep=""),"10")
axis(1,at=1:10,labels=xlabs)
ylabs <- formatC(seq(0,15000,2500),format="d",big.mark=",")
ylabs <- paste("$",ylabs,sep="")
axis(2,at=seq(0,15000,2500),ylabs,las=1)
mtext("Avg in-state tuition for 4-year public university",side=3)

legend(04,3000,c("Wyoming (Low)"),fill=c("#6E8B3D"),bty="n",cex=0.75)
legend(04,5700,c("Georgia"),fill=c("blue"),bty="n",cex=0.75)
legend(04,7300,c("US Avg"),fill=c("red"),bty="n",cex=0.75)
legend(04,12500,c("New Hampshire (Hi)"),fill=c("black"),bty="n",cex=0.75)

text(1,c(2500,4000,5500,9000),
     paste("$",formatC(tuition[1:4,1],format="d",big.mark=","),sep=""),pos=4,cex=0.7)

segments(1,c(2500,4000,5500,9000),1.1,c(2500,4000,5500,9000))
segments(1,2500,1,tuition[1,1])
segments(1,tuition[2:4,1],1,c(4000,5500,9000))


text(9,c(3500,7000,9000,13500),
     paste("$",formatC(tuition[1:4,10],format="d",big.mark=","),sep=""),pos=4,cex=0.7)

# This could be cleaned up using some programming approaches

segments(9.8,3500,10,3500)
segments(10,3500,10,tuition[1,10])

segments(9.8,7000,10,7000)
segments(10,7000,10,tuition[2,10])

segments(10,tuition[3,10],10,9000)
segments(10,9000,9.8,9000)

segments(9.9,13500,10,13500)
segments(10,13500,10,tuition[4,10])


```
```{r}
# Get one million vals from a uniform distro x <- runif(1000000,-3,3)
myhist <- function(pop,timestosamp, numtosamp, sleep=0.25) {
  avgvec <- vector() 
  length(avgvec) <- timestosamp
  for (ii in 1:timestosamp) {
  avgvec <- c(avgvec,mean(sample(pop,numtosamp))) 
  hist(avgvec, main=paste("Histogram - Iteration #",ii,sep=" "), xlim=c(-3,3))
  Sys.sleep(sleep)
  } 
}
myhist(x,500,5,sleep=.05)
```

