---
title: "BIOS 545 Lecture 4"
author: "Steve Pittard, wsp@emory.edu"
subtitle: Lists, Data Frames
output:
  pdf_document: default
  html_notebook: default
---

## Lists

Lists provide a way to store information of different types within a single data structure

Remember that vectors and matrices restrict us to only one data type at a time.
That is we cannot mix, for example, characters and numbers within a vector or matrix.

Many functions in R return information stored in lists

Consider the following example wherein we store information about a family. Not all this information is of the same type

```{r}
family1 <- list(husband="Fred", 
                wife="Wilma", 
                numofchildren=3, 
                agesofkids=c(8,11,14))

 length(family1)  # Has 4 elements
 
 family1
 
 str(family1)
```

If possible, always create named elements. It is easier for humans to index into a named list

```{r}
family1 <- list(husband="Fred", 
                wife="Wilma", 
                numofchildren=3, 
                agesofkids=c(8,11,14))
 # If the list elements have names then use "$" to access the element
family1$agesofkids
```

If the list elements have no names then you have to use numeric indexing

```{r}
(family2 <- list("Barney","Betty",2,c(4,6)))
```
```{r}
family2 <- list("Barney","Betty",2,c(4,6))

family2[4] # Accesses the 4th index and associated element

family2[[4]] # Accesses the 4th element value only - more direct

family2[3:4] # Get 3rd and 4th indices and associate values

```

As newcomers to R we usually doesn’t create lists except in two major cases:

1) We are writing a function that does some interesting stuff and we want to return to the user a structure that has information of varying types

2) As a precursor to creating a a data frame, which represents a hybrid object with characteristics of a list and a matrix

As an example of the first case, R has lots of statistical functions that return lists of information.


```{r}
data(mtcars) # Load mtcars into the environment 
mylm <- lm(mpg ~ wt + am, data = mtcars)
print(mylm)
```

But there is a lot more information

```{r}
str(mylm,attr=FALSE)
```

```{r}
names(mylm)

mylm$effects

lm(mpg ~ wt, data = mtcars)$coefficients
```

Some other basic R functions will return a list - such as some of the character functions:

```{r}
mystring <- "This is a test" 
mys <- strsplit(mystring, " ")
str(mys)

```

```{r}
mys[[1]][1]

mys[[1]][1:2]

unlist(mys)
```
When we create our own functions we can return a list

```{r}
my.summary <- function(x) { 
  return.list <- list() 
  return.list$mean <- mean(x) 
  return.list$sd <- sd(x) 
  return.list$var <- var(x) 
  return(return.list)
}
```


```{r}
my.summary(1:10)
```

Remember the sapply command ? We use it to apply a function over each element of a list or a vector. This helps us avoid having to write a "for-loop" every time we want to process a list or a vector.

```{r}
family1 <- list(husband="Fred", 
                wife="Wilma", 
                numofchildren=3, 
                agesofkids=c(8,11,14))

sapply(family1,class)
```


```{r}
sapply(family1,length)
```

sapply tries to return a "simplified" version of the output (either a vector or matrix) hence the ”s”in the ”sapply”. If you don’t use something like sapply then the example on the previous slide would look this:

```{r}
family1 <- list(husband="Fred", 
                wife="Wilma", 
                numofchildren=3, 
                agesofkids=c(8,11,14))

 for (ii in 1:length(family1)) {
    cat(names(family1)[ii]," : ",class(family1[[ii]]),"\n")
 }
```

Similar to sapply, the lapply function let’s you "apply" some function over each element of a list or vector. It will return a list version of the output hence the "l" in the lapply. So deciding between sapply and lapply simply is a question of format. What do you want back ? A vector or list ? Most of the time I use sapply.

```{r}
# lapply( vector_or_list, function_to_apply_to_each element)
family1 <- list(husband="Fred", 
                wife="Wilma", 
                numofchildren=3, 
                agesofkids=c(8,11,14))

lapply(family1,class)
```
check out the following:

```{r}
lapply(family1,mean)
```

We can write our own processing function that checks to see if the list element is valid input for the mean function.

```{r}
my.func <- function(x) { 
  if(class(x)=="numeric") {
    return(mean(x))
  }
}
lapply(family1, my.func)
```


See these videos on the lapply function at:

https://www.youtube.com/playlist?list=PL905DXZOAgwwj16m6C3ioh6aVKDDrEiiO

See this Blog post on lapply

https://rollingyours.wordpress.com/2014/10/20/the-lapply-command-101/


## Data Frames


Why should you use Data Frames ? 

- A data frame is a special type of list that contains data in a format that allows for easier manipulation, reshaping, and open-ended analysis

- Data frames are tightly coupled collections of variables. It is one of the more important constructs you will encounter when using R so learn all you can about it

- A data frame is an analogue to the Excel spreadsheet but is much more flexible for storing, manipulating, and analyzing data

- Data frames can be constructed from existing vectors, lists, or matrices. Many times they are created by reading in comma delimited files, (CSV files), using the read.table command

- Once you become accustomed to working with data frames, R becomes so much easier to use


Here we have 2 character vectors and 2 numeric vectors. Let’s say we want to do some summary on them:

```{r}
names <- c("P1","P2","P3","P4","P5") 
temp <-  c(98.2,101.3,97.2,100.2,98.5) 
pulse <- c(66,72,83,85,90)
gender <- c("M","F","M","M","F")
```


We could do some summary on this

```{r}
for (ii in 1:length(gender)) {
  print.string = c(names[ii],temp[ii],pulse[ii],gender[ii]) 
  print(print.string)
}

```
That doesn’t generalize at all. Use the dataframe() function to create a data frame. It looks like a matrix but allows for mixed data types


```{r}
names <- c("P1","P2","P3","P4","P5") 
temp <- c(98.2,101.3,97.2,100.2,98.5) 
pulse <- c(66,72,83,85,90)
gender <- c("M","F","M","M","F")

my_df <- data.frame(names,temp,pulse,gender) # Much more flexible

```

So now what ? 

```{r}
plot(my_df$pulse ~ my_df$temp,main="Pulse Rate",xlab="Patient",ylab="BPM")
     
mean(my_df[,2:3])
```

Once you have a data frame you could edit it with the Workspace viewer in RStudio although this doesn’t generalize. Imagine if your data set had 10,000 lines ?

### Builtin Example Data Frames

```{r}
library(help="datasets")
```

```{r}
help(mtcars)
data(mtcars)
str(mtcars)
```

Get some info on the data frame

```{r}
nrow(mtcars)
ncol(mtcars)
dim(mtcars)
```

More information is possible

```{r}
rownames(mtcars)
```

We can actually set the rownames using the same function

```{r}
rownames(mtcars) <- 1:32
head(mtcars)
```
```{r}
rownames(mtcars) <- paste("car",1:32,sep="_") 
head(mtcars)
```
There are many ways to index into or interrogate a data frame. This is where your previous knowledge of the bracket notation will be very useful to you

```{r}
head(mtcars[,-1])    # Get all rows / columns except for column 1
```

Compare this to:

```{r}
head(mtcars)
```
What about the following:

```{r}
mtcars[,] # same as
mtcars
```
```{r}
# Get first 5 rows and first 3 columns
mtcars[1:5,1:3] 
```
What about this ? 

```{r}
# Get all but the first 5 rows and first 3 columns
mtcars[-1:-5,-1:-3]
```

```{r}
# Get first 5 rows and the mpg and am columns
mtcars[1:5,c("mpg","am")]
```
The following is the same as above expcet we index into the columns using numbers instead of names. 

```{r}
mtcars[1:5,c(1,9)]
```
To find the names of all the columns

```{r}
names(mtcars)
```
So this is kind of boring because we can interrogate the data frame to find rows and columns that satisfy certain conditions

```{r}
# find all rows where the MPG >= 30
mtcars[mtcars$mpg >= 30,]
```


```{r}
# Find all rows where the mpg >= 30 but return only columns 2-6
mtcars[mtcars$mpg >= 30.0,2:6]

```

We can have compound statements. Find all rows where the mpg is >= 30 and the cylinder variable is less than 6. 

```{r}
mtcars[mtcars$mpg >= 30.0 & mtcars$cyl < 6,]
```

Find all rows that correspond to Automatic and count them. Frequently you just want to know how many rows satisfy as certain condition

```{r}
nrow(mtcars[mtcars$am == 0,])
nrow(mtcars[mtcars$am == 1,])

```
Of course there are other ways to do this

```{r}
table(mtcars$am)
```
So we can use other R functions as part of our interrogation query. Let's find all rows in the data frame where the MPG is greater than the mean MPG for the entire data set.

```{r}
mtcars[mtcars$mpg > mean(mtcars$mpg),]
```

Here is one which is slightly more involved. Let's find all the rows in the data frame where the MPG is greater than the 75th percentile for all MPG values in the data frame. To work this, you could break this down or do it all on one line.

```{r}
# Looks like the 4th element return represents the 75th percentile
quantile(mtcars$mpg)
```

```{r}
quantile(mtcars$mpg)[4]
```
```{r}
mtcars[mtcars$mpg > quantile(mtcars$mpg)[4],]
```
### Dealing with Factors

Factors are a special type of variable that we discussed last week. They can be identified by the fact that they ususally assume only a small number of unique values like < 10. 

```{r}
str(mtcars)
```
```{r}
unique(mtcars$am)
```
This begs the question - how many unique values does each column take on ? Is there a way to get that ?

```{r}
sapply(mtcars, function(x) length(unique(x)))
```

If we summarize one of these potential factors right now, R will treat it as being purely numeric which we might not want.

```{r}
summary(mtcars$am)
```
So this really isn’t helpful since we know that the ”am” values are transmission types

```{r}
mtcars$am <- factor(mtcars$am, levels = c(0,1), labels = c("Auto","Man") )
summary(mtcars$am)
```
We can add columns to a data frame. Let’s say we want to create a new column called“mpgrate”that, based on the output of the quantile command, will have a rating of the that car’s MPG in terms of "horrible","bad","good", or "great"

The labels could be more scientific but this is still a good use case. There are a couple of ways to do this:


```{r}
data(mtcars)   # Reload a "pure" copy of mtcars
mpgrate <- cut(mtcars$mpg,
               breaks = quantile(mtcars$mpg),
               labels=c("horrible","Bad","Good","Great"),include.lowest=T) 
mtcars <- cbind(mtcars,mpgrate)
head(mtcars[,10:12])
```
We could also add a new column like:

```{r}
mtcars$mpgrate <- mpgrate
```

Why go to all this trouble ? Because then the plotting functions know how to deal with the data more easily.

```{r}
library(lattice) 
bwplot(~mpg|mpgrate,data=mtcars,layout=c(1,4))

```

Transforming columns is a common activity. 

```{r}
transform(mtcars,wt = (wt*1000), 
                 qsec = round(qsec),
                 am = factor(am,labels=c("A","M")))
```

### Missing Values

The NA (datum Not Available) is R’s way of dealing with missing data. NAs can give you trouble unless you explicitly tell functions to ignore them.
You can also pass the data through na.omit(), na.exclude(), or complete.cases() to insure that R handles data accordingly.

```{r}
data <- data.frame(x=c(1,2,3,4), 
                   y=c(5, NA, 8,3),
                   z=c("F","M","F","M"))

data
```

```{r}
na.omit(data)
```

```{r}
data <- data.frame(x=c(1,2,3,4), 
                   y=c(5, NA, 8,3), 
                   z=c("F","M","F","M"))
complete.cases(data)

```

```{r}
sum(complete.cases(data)) # total number of complete cases 

sum(!complete.cases(data)) # total number of incomplete cases 

data[complete.cases(data),] # Same as na.omit(data) 
```

```{r}
url <- "https://raw.githubusercontent.com/steviep42/bios545_spring_2021/master/DATA.DIR/hs0.csv"

data1 <- read.table(url,header=F,sep=",")

names(data1) <- c("gender","id","race","ses","schtyp","prgtype",
                   "read", "write","math","science","socst")
```


```{r}
head(data1)

```

```{r}
nrow(data1)

sum(complete.cases(data1)) 
sum(!complete.cases(data1)) 
data1[!complete.cases(data1),]
```

Many R functions have an argument to exclude missing values

```{r}
data1[!complete.cases(data1),]
```

```{r}
mean(data1$science)

mean(data1$science,na.rm=T)
```

Many times data will be read in from a comma delimited ,("CSV"), file exported from Excel. The file can be read from local storage or from the Web.

```{r}
url <- "https://raw.githubusercontent.com/steviep42/bios545_spring_2021/master/DATA.DIR/hsb2.csv"

data1 <- read.table(url,header=T,sep=",")
head(data1)
```

Let's look at a "real" file. I got a file from this site https://data.cityofchicago.org/ and put it on a server if you want to download it and give it a whirl.

Also, my laptop has 8GB of RAM. I suspect if you have 2GB of RAM on your laptop you will be okay but I cannot be sure. On campus it took about 30 seconds to download and read it into R.

I zipped it to make it more manageable. You can download it and read it into R using the following commands

```{r}
url <- "https://github.com/steviep42/bios545_spring_2021/blob/master/DATA.DIR/chi_crimes.csv.zip?raw=true"

download.file(url,destfile="chi_crimes.csv.zip")

chi <- read.csv(unzip("chi_crimes.csv.zip"),
                  header=TRUE,sep=",",stringsAsFactors = FALSE)
```



```{r}
names(chi)
```

```{r}
sapply(chi, function(x) length(unique(x)))
```

```{r}
# Make the date a "real date"
chi$Date <- strptime(chi$Date,"%m/%d/%Y %r")
chi$month <- months(chi$Date)
chi$month <- factor(chi$month,levels=c("January","February","March",
                    "April","May","June","July","August", "September",
                    "October","November","December"),ordered=TRUE)

```


```{r}
# Okay how many crimes were committed in each Month of the year ?
plot(1:12,as.vector(table(chi$month)),
     type="n",xaxt="n", ylab="Alleged Crimes",
     xlab="Month", main="Chicago Crimes in 2012 by Month",ylim=c(5000,33000)) 
grid()

axis(1,
     at=1:12,
     labels=as.character(sapply(levels(chi$month), function(x) substr(x,1,3))),cex.axis=0.8)

points(1:12,as.vector(table(chi$month)),type="b",pch=19,col="blue")
points(1:12,as.vector(table(chi$month,chi$Arrest)[,2]),
       col="red", pch=19,type="b")
```


```{r}
barplot(table(chi$Arrest,chi$month),
        col=c("blue","red"),
        cex.names=0.5, 
        main="Chicago: Reported Crimes vs. Actual Arrests")

legend("topright",c("Arrests"),fill="red")
 # Even easier to do
hold <- rev(sort(table(chi$month)))
barplot(rev(sort(hold)),horiz=F,
        las=1,cex.names=0.5,col=heat.colors(12), 
        main="Chicago: Reported Crimes in 2012 by Month")
```



```{r}
# Find out number of alleged crimes by type
categories <- rev(sort(sapply(unique(as.character(chi$Primary.Type)), function(x) { nrow(chi[chi$Primary.Type==x,]) })))

categories <- rev(sort(table(chi$Primary.Type)))

barplot(categories,horiz=F,las=1,cex.names=0.6,col=heat.colors(30), las=2, main="Chicago: Types of Crimes Reported")
```


```{r}
hold <- chi[chi$Primary.Type == "GAMBLING",]
hold <- chi[chi$Primary.Type == "GAMBLING" & chi$Description != "GAME/DICE",]
nrow(hold) # How many non-Dice related gambling offenses were there ? # About 26 I think
# Let's plot them on a map
library(googleVis) # This is an addon package you must install
hold$LatLon <- paste(hold$Latitude,hold$Longitude,sep=":")
hold$Tip <- paste(hold$Description,hold$Locate.Description,hold$Block,
"<BR>",sep=" ") 

chi.plot <- gvisMap(hold,"LatLon","Tip")
plot(chi.plot)
```


## Review apply, sapp,y, and lists

```{r}
set.seed(1)  # Makes the call to rnorm generate the same numbers every time
( mymat <- matrix(round(rnorm(16,10),2),4,4) )
```

Let’s say that we want to take the mean of each column. Another way to say this is that we want to apply the mean function to each column in the matrix. This is the way to start thinking about things like this in R. But first, how might we do this if we don’t know anything about the apply command ?

```{r}
mean(mymat[,1])

 
mean(mymat[,2])

 
mean(mymat[,3])

 
mean(mymat[,4])
```

We could put this into a vector

```{r}
( mymatcolmean <- c(mean(mymat[,1]),mean(mymat[,2]),mean(mymat[,3]),mean(mymat[,4])) )
```

But we could easily do this outselves using the apply function

```{r}
apply(mymat, 2, mean)

apply(mymat, 1, mean)
```

And we could use functions other than **mean** such as getting the sums of the columns

```{r}
apply(mymat,2,sum)    # Get the sum of all the columns
```

Imagine if the matrix was really big. Here is a matrix with 100,000 elements from a normal 
distribution

```{r}
set.seed(123)
bigmat <- matrix(rnorm(100000),nrow=1000,ncol=100)
```

Now let's get the means of all the columns

```{r}
apply(bigmat,2,mean)
```

Check it out to make sure it works. According to this, the mean of column one is 0.0161278659

```{r}
mean(bigmat[,1])
```


Or get the max value from each column

```{r}
apply(bigmat,2,max)
```

We could also pass arguemnts to any R functions we might use. For example, what if we wanted to trim the top and botoom percent of the data before taking the mean?

```{r}
apply(bigmat,2,mean,trim=0.05)
```

This is every efficient. Let's get find the max value of each row of a 1,000,000
element matrix and then from that, get the over all max calue

```{r}
set.seed(234)
biggermat <- matrix(rnorm(1000000),1000,1000)
```


```{r}
# Get the max value of each row
maxrows <- apply(biggermat,1,max)

# Get the max of that
max(maxrows)

```

Or maybe just do this. 

```{r}
max(apply(biggermat,1,max))
```

```{r}
system.time(max(apply(biggermat,1,max)))
```

Remember we have lists that frequently come back from many R stat functions.

```{r}
x <- 1:10             # x vals
y <- 1:10 + rnorm(10,2,2)   # A y value plus some noise
mylm <- lm(y~x,data.frame(x,y))
```

Let's figure out some things about what we get back
```{r}
str(mylm,0)

str(mylm,1)
```


We also encountered lists when using the strsplit function

```{r}
somestring <- "Hello. My name is John. Your name is Mary."

(mys <- strsplit(somestring," "))


```

We could reverse this string using the lapply function
```{r}
lapply(mys,rev)
```

