---
title: "BIOS 545 Graphics Part 2"
output:
  html_document:
    df_print: paged
---

### Colors Revisited

We left off with a discussion on colors and how they can improve a chart or plot. We found that we could use built-in palettes from which we could pick "pre- made" colors. 

This makes it easy for those of us who can't really pick matching color schemes so easily.

This week we extend this idea some and learn about "color ramps", annotation, as well as some other chart types.

```{r}
myt <- tapply(mtcars$mpg, mtcars$cyl, mean) # tapply produces a table
barplot(myt, main = "Mean MPG per Cylinder Group", col=heat.colors(3))

```

```{r}
myt <- tapply(mtcars$mpg, mtcars$carb, mean)
barplot(myt, main = "Mean MPG per Carb group",
        col= terrain.colors(length(myt)),legend=TRUE)
```

```{r}
barplot(rev(sort(myt)),
        main = "Mean MPG per Carb group",
        col= terrain.colors(length(myt)),
        legend=T)
```
```{r}
barplot(rev(sort(myt)), 
        main = "Mean MPG per Carb group", 
        col= terrain.colors(length(myt)),
        legend=T)
sort(myt)
```

What about making the colors reflect the mean ? If we have an increasing sequence of bars why not make the color for each group a different shade from a graduated color scale.

To do this we need a "color ramp" command to generate a palette. Its easier to understand this with an example.

This creates a graduated color scheme between red and blue. Let's apply this to our barplot.

```{r}
mycols <- colorRampPalette(c("yellow","blue"))( length(myt))
mycols
```

```{r}
barplot(rev(sort(myt)),
        main = "Mean MPG per Carb group", 
        col=mycols, legend=T)
```

There are other color palette generators available


```{r}
# Run display.brewer.all() to see all palettes
library(RColorBrewer)
mypalette <- brewer.pal(length(myt),"Spectral")

barplot(rev(sort(myt)),
        main = "Mean MPG per Carb group", 
        col=mypalette, legend=T)
```

### Annotating plots

Many Base graphics functions in R will return useful information that can then be used to annotate the graph. Let's generate a table which summarizes the mean MPG per cylinder group.


```{r}
(myt <- tapply(mtcars$mpg, mtcars$cyl, mean))
```

```{r}
temp <- barplot(myt, 
                main = "Mean MPG per Cylinder Group",
                col=heat.colors(3))

temp
```

So what is this information contained in **temp** ? These values represent the X coordinates for each bar.

```{r}
temp
```

```{r}
temp <- barplot(myt, 
                main = "Mean MPG per Cylinder Group",
                col=heat.colors(3))


# We put up the means one by one

text(0.2, 25, "26.66")
text(1.9, 10, "19.74")
text(3.1, 10, "15.10")
```
But this is working too hard

```{r}
n=10
temp <- barplot(myt, 
                main = "Mean MPG per Cylinder Group",
                col=heat.colors(3))
text(temp,n,round(myt,2))
```
### Boxplots

If you are doing a X/Y scatterplot and one of them is a categorical variable then the resulting plot will be a boxplot. Of course, you could always plot a single continuous variable as a boxplot.

A box plot summarizes a lot of information clearly. It shows the location and spread of data as well as skewness. The horizontal line shows the median. The bottom and top lines show the 25th and 75th percentiles respectively.

The vertical dashed lines are called "whiskers". The show the maximum of 1) the smaller of the data being plotted or 1.5 times the interquartile range.

1.5 times the interquartile is roughly two standard deviations. And the IQR is the 
difference in the response variable between the first and third quartile.

Note that any boxplot output will also match the output from the "fivenum" command:

```{r}
fivenum(mtcars$mpg)
boxplot(mtcars$mpg, main="MPG Boxplot",xlab="MPG",col="aquamarine")

# try working with the log of the variable
boxplot(log(mtcars$mpg), main="MPG Boxplot",xlab="MPG",col="aquamarine")

```

We can get more than one boxplot within a plot window. If you are doing a X/Y scatterplot and one of them is a categorical variable then the resulting plot will be a boxplot. Let's look at the mtcars dataset now.

```{r}
boxplot(mpg~am, 
        data = mtcars, 
        main="MPG / Transmission",
        col=heat.colors(2), names=c("auto","manual"))

legend(0.5,33,c("Auto","Manual"),fill=heat.colors(2))
```

```{r}
boxplot(mpg~cyl,data=mtcars,
        main="MPG by Cylinders", 
        col=rainbow(3), 
        horizontal=TRUE, names=c("Four","Six","Eight"))

legend("topright",c("4","6","8"),fill=rainbow(3))
```
Boxplots are generally good at showing distribution of data around the median but not so good at showing the significance of differences between medians. Tukey introduced the idea of "notched" plots to address this problem.

Boxes for which the notches do not overlap are "likely" to have significantly different medians in terms of testing.

The width of the notches is proportional to the interquartile range of the sample and inversely proportional to the square root of the size of the sample.

```{r}
boxplot(mpg~cyl, 
        main="MPG vs. Cylinder", 
        data=mtcars, notch=TRUE, col=terrain.colors(3))
```

Just as with the bar charts, we can annotate / draw on top of an existing boxplot. We just have to understand the information being returned by the boxplot function.

```{r}
hold <- boxplot(mtcars$mpg, main="MPG Boxplot", xlab="MPG",col="aquamarine")
```

```{r}
hold
```

There is only one boxplot so the X coordinate is "1".

```{r}
hold <- boxplot(mtcars$mpg, main="MPG Boxplot", xlab="MPG",col="aquamarine")
pos = 3
text(1,10.4,10.4,pos=pos) 
text(1,15.35,15.35,pos=pos) 
text(1,19.20,19.20,pos=pos) 
text(1,22.8,22.8,pos=pos) 
text(1,33.9,33.9,pos=1)    # This is so the label won't be in the top margin
```
We could do this more simply by recognizing that the text function can handle vectors as well as single values (as in the previous example).

```{r}
hold <- boxplot(mtcars$mpg, main="MPG Boxplot", xlab="MPG",col="aquamarine")
text(1,hold$stats[,1],hold$stats[,1],pos=1) 
```

So, in the case of multiple boxplots, we can get the same type of information:

```{r}
tmpvar <- boxplot(mpg~cyl,data=mtcars)
```

```{r}
tmpvar
```

```{r}
tmpvar <- boxplot(mpg~cyl,data=mtcars)
tmpvar$stats
     
tmpvar$stats[3,] 
mtext(at=1:3, text=tmpvar$stats[3,])
```


We could also do this:

```{r}
tmpvar <- boxplot(mpg~cyl,data=mtcars)
tmpvar$stats
     
tmpvar$stats[3,] 

mtext(at=1, text=tmpvar$stats[3,1], side=3)
mtext(at=2, text=tmpvar$stats[3,2], side=3)
mtext(at=3, text=tmpvar$stats[3,3], side=3)
```

### More Experimentations with Color and Annotation

Let's look at a time series example involving some real data. This is motivated by some data which comes from National Morbidity and Mortality Air Pollution Study (NMMAPS).


```{r}
url <- "https://raw.githubusercontent.com/steviep42/bios545_spring_2021/master/DATA.DIR/chicago-nmmaps.csv"
nm <- read.csv(url,as.is=TRUE)

str(nm)

sapply(nm,function(x) length(unique(x)))
```


```{r}
# We need to convert the date strings into actual Dates
nm$date <- as.Date(nm$date)

# We pull out the records after 12/31/96
nm <- nm[nm$date > as.Date("1996-12-31"),]

# All we care about is the year in XXXX format
nm$year <- substring(nm$date,1,4)

# So we plot it
plot(temp~date,data=nm,pch=18,cex=0.7,col="maroon")
grid()
```


If we want to use different colors for each of the four seasons we have to do this manually. That is, we:

1. have to pick the colors ourselves
2. setup a null plot
3. use the split command to partition the data frame into 4 groups which correspond to each of the four seasons. 
4. then loop through the four "splits" and use the *points* function to draw the points for a given season while indexing into the color vector. 
5. Create the legend and the axes limits


```{r}
library(RColorBrewer)

# Need to make some extra room for the legend
ylim <- c(min(nm$temp)-5,max(nm$temp)+5)

plot(temp~date,
     data=nm,
     pch=18,
     cex=0.7,
     type="n",
     ylim=ylim)

mycols <- brewer.pal(4,"Set2")

# Split the nm data frame on season labels
splits <- split(nm,nm$season)

str(splits,1)   # The str() function let's us peek at the structure

# Loop through the splits and use the points function to 
# plot the temperature per season 

for (ii in 1:length(mycols)) {
  points(splits[[ii]]$date,
         splits[[ii]]$temp,
         col=mycols[ii],
         pch=18,
         cex=0.9)
}

title(main="Temperature Across Seasons")
legend("topleft",
       names(splits),
       pch=18,
       col=mycols,
       horiz=TRUE,
       cex=0.9,
       pt.cex=1.4)
grid()
```


That was a lot of work. Or was it ? 

At a minimum you have to know something about R programming to get this done. Of course having knowledge of programming techniques is never a bad thing although, for beginners, when one is visualizing data it is usually best to use tools that don't require alot of effort to view the data in interesting ways.

Each chart type has it's own function with it's own arguments that influence that particular chart outcome. You have to dig into the help pages to figure out how to do things such as putting text on the chart


```{r}
r <- hist(nm$temp,
          col="aquamarine",
          breaks=12,
          main="Temperature Degrees F",
          xlab="Temperature")

text(r$mids, r$density, r$counts, adj = c(.5, -.5), col = "black")
# or
text(r$mids, r$counts, r$counts, adj = c(.5, -.5), col = "black")

```

If we wanted to look at temperatures as "conditioned"" by season we need to use the *par* function to carve out 4 panels. We try to plot such that the axes for all plots are the same to enable better comparisons. 

```{r}
par(mfrow=c(2,2))  # Get 2 rows by 2 columns

xlim <- range(nm$temp)
xlab <- "Temperature"
ylim <- c(0,140)
col <- "lightblue"

# Get the season names
seasons <- unique(nm$season)
tempstr <- "Temperature in"

for (ii in 1:length(seasons)) {
  title  <- paste(tempstr,seasons[ii],sep=" ")
  hist(nm[nm$season==seasons[ii],]$temp, 
       main=title,
       xlab=xlab,
       ylim=ylim,
       col=col,
       xlim=xlim)
}
par(mfrow=c(1,1)) # Reset plot window to 1 row by 1 column

```
it might be easier to look at the temp distribution per season as boxplots within a single plot window:

```{r}
title <- "Boxplots Across Seasons"
boxplot(temp~season,
        data=nm,
        notch=TRUE,
        col="lightblue",
        main=title)
```


Multiple Graphics Packages

We are focusing on 3 graphics packages (well 2 actually)

1. Base Graphics - Oldest and Most Well Known
2. lattice - Second Oldest and Supports Grouping and Panels
3. ggplot - Third Oldest and Supports Grouping and facets (similar to Panels)

```{r}
data(mtcars)
xlab <- "Wt in Lbs/1,000"
main <- "MPG vs Wt"
plot(mpg~wt,data=mtcars,main=main,xlab=xlab)
grid()
```

Lattice was written to provide grouping and paneling

- Consistent look and feel
- Great for multivariate data
- Takes care of lots of things for you
- Has a formula interface 
- Lots of examples and support on Google
- See \url{http://lmdvr.r-forge.r-project.org/figures/figures.html} 
- Picks useful defaults for you


Lattice, like Base Graphics, work on the idea that you locate the function of interest to do the plotting. Consequently there are multiple functions corresponding to a given plot top. Thankfully, lattice did NOT name their functions to be the same as the Base Graphics functions else there would have been a real problem. Here are some of the functions as well as their Base graphics equivalent

Lattice Function | Description   | Base Graphics Equivalent
-----------------|---------------|-------------------------
barchart()       | Barcharts     | barplot()
bwplot()         | Boxplots      | boxplot()
doptlot()        | Dotplots      | dotchart()
histogram()      | Histograms    | hist()
stripplot()      | Strip Plots   | stripchart()
xyplot()         | Scatter plot  | plot()
qq        | Quantile-quantile plots | qqplot()
qqmath() | Quantile-quantile plots Data set vs Data set| qqplot() 
splom()          | Scatterplot matrices| pairs()
levelplot() | Level Plots | image()
contourplot() | Contour Plots | contour()



```{r}
library(lattice)
xlab <- "Wt Lbs / 1,000"
main <- "MPG vs. Wt"
xyplot(mpg~wt,data=mtcars,main=main,xlab=xlab,type=c("p","g"))

```


```{r}
library(lattice)
xlab <- "Wt Lbs / 1,000"; main <- "MPG vs. Wt"
xyplot(mpg~wt,
       groups=factor(cyl),
       data=mtcars,
       main=main,
       xlab=xlab,
       type=c("p","g"),auto.key=list(columns=3))

```

```{r}
library(lattice)
xlab <- "Wt Lbs / 1,000"; main <- "MPG vs. Wt"
xyplot(mpg~wt|factor(cyl),data=mtcars,main=main,xlab=xlab,
       type=c("p","g"),layout=c(3,1))
```

We can use colors to distinguish between transmission type within these panels.

```{r}
xyplot(mpg~wt|factor(cyl), data=mtcars, groups=factor(am), pch=19, 
       cex=1.3,layout=c(3,1),type=c("p","g"),main="MPG vs Wt per Cylinder Group",
       auto.key=TRUE)
```

Doing Panels in Base Graphics is a Manual Process. This can be Powerful but very tedious

```{r}
xlab <- "Wt Lbs / 1,000"; main <- "MPG vs. Wt"
par(mfrow=c(1,3))
maxmpg <- max(mtcars$mpg)
maxwt  <- max(mtcars$wt)
mydf <- split(mtcars,mtcars$cyl)
for (ii in 1:length(mydf)) {
  tmpdf <- mydf[[ii]]
  main <- paste("MPG vs. Wt",names(mydf)[ii],sep=" - ")
  plot(mpg~wt,data=tmpdf,main=main,
                         xlim=c(0,maxwt),
                         ylim=c(0,maxmpg))
  grid()
}
```
So remember above when we created a 2 x 2, conditioned plot of the temperature by the seasons ? We can replace that in one command in lattice.

```{r}
histogram(~temp|season,data=nm)
```


### Modes 

Independently of the graphics packages we operate in two modes: Exploratory and Publication

- Publication mode is for, well, publication and sharing
- This is for journals
- Graphs are usually standard types but combined in interesting ways
- Comparisons across panels and facets
- Axes, legends, titles, colors, groups are usually present
- ggplot makes it easy to create good default graphics

### ggplot2

ggolpt2 is a relative newcomer to R

- Written according to a ''Grammar of Graphics'' (Wilkinson, 2005)

- Discuss the visualization of data using an accepted vocabulary

- Flexibly explore the data using a number of visualizations (or combinations thereof)

- No need to ''commit'' to specific chart types

- Specify plots in abstract terms using the aforementioned grammar

- Rapidly becoming the default R graphics package

- Attempts to leverage the good parts of lattice and Base graphics


Some helpful resources:

- Book - [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/)
- [Vanderbilt Workshop](http://ggplot2.org/resources/2007-vanderbilt.pdf)
- [Mailing List](http://groups.google.com/forum/?fromgroups#!forum/ggplot2)
- [Documentation](http://ggplot2.tidyverse.org/reference/)
- [R for Data Science](http://r4ds.had.co.nz/) Online Book 
- [Cheat Sheet](http://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) 
- [R Graphics Cookbook](http://www.cookbook-r.com/Graphs/index.html)


ggplot2 is part of the ''tidyverse'' which is a collection of R packages that share common philosophies to work well together

- Home page for project is {http://tidyverse.org/
- Main packages are: ggplot2, tibblr, tidyr, readr, purrr, and dplyr
- Can install from within R Studio just like any other package
- The name of the package is simply **tidyverse**


#### qplot

**qplot** is a command in ggplot that offers a ''training wheels'' like capability:

- The ''q'' in **qplot** stands for ''quick''
- Is supposed to be an analog of the Base graphics **plot** command
- Useful if you already know Base Graphics really well
- Convenient wrapper for quickly creating a number of different types of plots
- However it's not really demonstrative of the power of grammar of graphics 

```{r}
library(ggplot2)
# A quick plot
qplot(mpg,wt,data=mtcars,main="MPG vs wt")

# Group using colors
mtcars$am <- factor(mtcars$am,labels=c("Auto","Manual"))
qplot(mpg,wt,data=mtcars,main="MPG vs wt",col=am)
qplot(mpg,wt,data=mtcars,main="MPG vs wt",facets=am ~ .)

```

Look at a paneled plot for each value of cylinder

```{r}
qplot(y=mpg,x=wt,data=mtcars,main="MPG vs wt",facets=cyl ~ .)

```

Specifying a "geometry" is a powerful part of ggplot

```{r}
qplot(mtcars$mpg, geom = "histogram", bins = 12, color = I("black"), fill=I("white"))
```

We can easily put a smoothing line over the data
```{r}
qplot(wt, mpg, geom = "point", data=mtcars) + geom_rug() + geom_smooth()
```

### Key Ideas

There are some essential terms and concepts for using ggplot. These aren't the only ones but they are some of the most important:

- Data: The actual data frame under consideration
- Aesthetics: Visual elements mapped to the data (axis, lines, colors, bars, etc)
- Scales: Transformations you might want to apply (e.g. logarithm, polar coordinates)
- Geometries: The shape mapped to the aesthetic(s)

#### Aesthetics

Here are some of the aesthetics that help make a plot:

- x and y position
- size of the elements
- shape
- color

#### Geometries

We use geometries to view the data:

- lines and variations (dashed, segments, etc)
- bars, histograms
- text labels
- points

See http://ggplot2.tidyverse.org/reference/#section-layer-geoms


### mtcars

We’ll first use the built-in mtcars data frame to explore these ideas since mtcars is an easy data frame to understand:

The data was extracted from the 1974 Motor Trend US magazine,
and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).


```{r}
data(mtcars)
head(mtcars)
```


What are the categories/factors in this data ? What are the continuous quantities ?

```{r}
str(mtcars)
```


What are the categories/factors in this data ?
What are the continuous quantities ?

```{r}
sapply(mtcars, function(x) length(unique(x)))
```

Variables taking on a limited number of unique values are probably factors

Variables taking on many different unique values are probably continuous

We generally seek to summarize and/or compare continuous information in terms of (or across) categories
 
#### Questions

Back to the data, Some driving questions might be:

- What does the distribution of wt values look like ?

- Is there a relationship between mpg vs wt ?

- Does mpg appear to be different over individual cylinder groups ?

- Is there a relationship between mpg and wt that is effected by cylinder group ?

- Does mpg appear to be different over different transmission types ? 

- What are the counts of transmission types and cylinder groups ?

What does the distribution of wt values look like ?

```{r}
ggplot(mtcars,aes(x=wt)) + 
  geom_histogram(bins=15) + 
  ggtitle("Steve's plot")
```
What if we don't supply a geometry ?

```{r}
ggplot(mtcars,aes(x=wt))
```
Is there a relationship between mpg vs wt ?

```{r}
ggplot(mtcars,aes(x=wt)) + geom_point(aes(y=mpg))
```
Is there a relationship between mpg vs wt ?

Note how we added a new geometry on an existing aesthetic mapping then added another aesthetic mapping - we mapped the y-axis to the mpg variable.

```{r}
ggplot(mtcars,aes(x=wt)) + geom_point(aes(y=mpg))

# Could have also started over - see how flexible ggplot can be

ggplot(mtcars,aes(x=wt,y=mpg)) + geom_point()
```


Is there a relationship between mpg vs wt ? We can add titles, labels, and captions

```{r}
ggplot(mtcars,aes(x=wt,y=mpg)) + 
  geom_point() + 
  ggtitle("MPG vs Wt","mtcars data frame") + 
  labs(caption="Extracted from 1974 Motor Trend US")
```

We can add another geometry. In this case a regression line with confidence intervals:

```{r}
ggplot(mtcars,aes(x=wt,y=mpg)) + 
  geom_point() + 
  ggtitle("MPG vs Wt","mtcars data frame") +
  geom_smooth(method="lm")
```

Question: Does mpg appear to be different over individual cylinder groups ?

Before answering this question I think that it is a good time to introduce the idea of ”grouping”:

- For each unique value of a factor or category we can see how it impacts the plot
- We can use color, shapes, and size to accomplish this
- In ggplot we use and“aesthetic mapping”to do this
- Note that the cyl variable assumes only three unique values

```{r}
unique(mtcars$cyl)
```
Let's make cyl and "official" factor

```{r}
mtcars$cyl <- factor(mtcars$cyl)
ggplot(mtcars,aes(x=wt,y=mpg,color=cyl)) + 
  geom_point() +
  ggtitle("MPG vs Wt")
```
What happens if we use a continuous quantity as a color aesthetic ? Like hp ?

```{r}
ggplot(mtcars,aes(x=wt,y=mpg,color=hp)) + geom_point()
```

We can use multiple variables for grouping

```{r}
ggplot(mtcars,aes(x=wt,y=mpg,size=hp,color=cyl)) + geom_point()
```
Maybe better to put aesthetic assignments in the associated geom layer. This gives us flexibility

```{r}
ggplot(mtcars) + geom_point(aes(x=wt,y=mpg,size=hp,color=cyl))
```


- Note that there is a difference between mappings and settings with the former usually being a function of some variable in the data

- Settings are for altering appearance in a fixed, "set" way

```{r}
library(ggplot2)
library(gridExtra)  # you might have to install this

p1 <- ggplot(mtcars,aes(x=wt,y=mpg)) + geom_point(color="red")

p2 <- ggplot(mtcars,aes(x=wt,y=mpg)) + geom_point(color="red",size=4) 

grid.arrange(p1, p2, nrow=1, ncol=2)
```

In general anything you wish to set to a static value should be set outside of the aes function. Here we create a density plot of MPG. Note with a static color assignment we place it outside of the aes

```{r}
ggplot(mtcars) + geom_density(aes(x=mpg),fill="aquamarine")
```

Here we create a density plot of MPG. We can group the density by cylinder group. Notice too, how I can specify a theme to influence the appearance of the background plot area. 

```{r}
ggplot(mtcars) + geom_histogram(aes(x=mpg,fill=cyl)) + theme_bw()
```

What about looking at a boxplot of mpg across cylinder groups ?

```{r}
ggplot(mtcars) + geom_boxplot(aes(x=cyl,y=mpg)) 
```

Let's look at some wage data to help us better understand the ggplot package

```{r}
url <- "https://raw.githubusercontent.com/pittardsp/bios545r_spring_2018/master/SUPPORT/wage.csv"
wages <- read.csv(url) 

str(wages)
```

```{r}
ggplot(wages,aes(x=age,y=wage)) + geom_point() + ggtitle("Wage vs Age")
```
```{r}
unique(wages$jobclass)
ggplot(wages,aes(x=age,y=wage,color=jobclass)) + 
  geom_point() + ggtitle("Wage vs Age") +
  theme_bw()
```

```{r}
ggplot(wages,aes(x=age,y=wage,color=sex)) + 
  geom_point() + ggtitle("Wage vs Age") +
  theme_bw()
```
```{r}
ggplot(wages,aes(x=age,y=wage,color=education)) + 
  geom_point() + ggtitle("Wage vs Age") +
  theme_bw()
```



```{r}
ggplot(wages,aes(x=age,y=wage,color=maritl)) + 
  geom_point() + ggtitle("Wage vs Age") +
  theme_bw()
```


Facets allow us to display data within panels to facilitate easy comparisons In lattice graphics this is called paneling or conditioning.

In Base graphics you have to do paneling or faceting manually - it’s a pain
Consider the following grouping graph:

```{r}
ggplot(wages,aes(x=age,y=wage,color=maritl,alpha=0.5)) + 
  geom_point() + ggtitle("Wage vs Age") + facet_wrap(~education) +
  theme_bw() + guides(alpha=FALSE)
```

```{r}
ggplot(wages,aes(x=age,y=wage,color=maritl,alpha=0.5)) + 
  geom_point() + ggtitle("Wage vs Age") + facet_grid(~education) +
  theme_bw() + guides(alpha=FALSE)
```
```{r}
ggplot(wages,aes(x=age,y=wage,color=jobclass)) + 
  geom_point() + ggtitle("Wage vs Age") + facet_grid(health~education) +
   guides(alpha=FALSE) + scale_colour_viridis_d(option = "plasma")
```

Let's get back to mtcars to illustrate some more concepts and answer some questions

- plot mpg vs wt for all combinations of cylinder group and Transmission Type

- Let’s make these variables into factors 

- What combination gets the best MPG ?

```{r}
mtcars <- transform(mtcars,am=factor(am,labels=c("A","M")), cyl=factor(cyl))

ggplot(mtcars,aes(x=wt,y=mpg)) + 
  geom_point() + 
  facet_wrap(~cyl+am) + ggtitle("MPG vs Wt") + xlab("Wt in Lbs/1,000")
```


```{r}
mtcars$am <- factor(mtcars$am,labels=c("auto","manual"))

ggplot(mtcars,aes(x=wt,y=mpg,color=am)) + 
       geom_point() + 
       ggtitle("MPG vs. Weight by Cylinder Group") +
       xlab("Weight / 1,000") +
       ylab("Miles Per Gallon") + 
       theme_light() +
       facet_wrap(~cyl) 
```

### Count and Tabular Data

What about tabular data ?

- ggplot usually prefers data frames as input
- aggregation commands in R produce tables
- If you have the dataframe used to produce the table then use that

```{r eval=FALSE}
# This won't work
myt <- table(mtcars$am)
ggplot(myt,aes(x=am)) + geom_bar() +
                           ggtitle("Distribution of Transmission Types")
```
```{r}
mtcars$am <- factor(mtcars$am, labels=c("Auto","Manual")) 
ggplot(mtcars,aes(x=am)) + 
  geom_bar() +
  ggtitle("Distribution of Transmission Types")
```
```{r}
ggplot(mtcars,aes(x=factor(carb))) + 
  geom_bar() +
  ggtitle("Observations by Carburetor")
```

But What if you are starting with a table ?

```{r}
(ctab <- table(carb=mtcars$carb))

# Remember that ggplot wants a data frame
(df <- as.data.frame(ctab))

ggplot(df,aes(x=carb,y=Freq)) + geom_bar(stat="identity") +
                                     ggtitle("Observations by Carburetor")
```


How do we order the bars ? Remember that ggplot wants a data frame. We also use a "stat" of
"identity" since we are using a pre-existing table that already has # count information.
geom_bar uses "stat_count" by default which does the counting for you. But we already have the counts in our table.

```{r}
(df <- as.data.frame(ctab))

ggplot(df,aes(x=reorder(carb,Freq),y=Freq)) + 
  geom_bar(stat="identity") + ggtitle("Observations by Carburetor")

```


- If you want to look at a 2-way count table then you can use the fill aesthetic.

- This is one of those things that actually winds up being easier in Base graphics

- Although once you learn the “grammar of graphics”this becomes easier

- This relates mostly to table and count data

```{r}
ggplot(mtcars,aes(x=am)) + 
  geom_bar(aes(fill=cyl)) + 
  ggtitle("Tranmission by Cylinder Group") +
  xlab("Transmission Type") +
  ylab("Count by Cylinder Group")
```


```{r}
mtcars <- transform(mtcars,am=factor(am,labels=c("Auto","Manual")),cyl=factor(cyl))
ggplot(mtcars,aes(x=am)) + 
  geom_bar(aes(fill=cyl),position="dodge") + 
  ggtitle("Tranmission by Cylinder Group") +
  xlab("Transmission Type") + 
  ylab("Count by Cylinder Group - Side by Side")
```
But what if you have just the table without data from whence it came ?

```{r}
sometable <- table(transmission=mtcars$am,cyl=mtcars$cyl)
df <- as.data.frame(sometable)

ggplot(df,aes(transmission,Freq,fill=cyl)) + geom_bar(stat="identity") +
                           ggtitle("Tranmission by Cylinder Group") +
                           xlab("Transmission Type") +
                           ylab("Count by Cylinder Group")
```
Here is how to list the cars in mtcars by MPG from best to worst


```{r}
# Make Cylinder into a factor
mtcars$cyl <- factor(mtcars$cyl,labels=seq(4,8,2))

# Use reorder to arrange the data from high MPG to low
ggplot(mtcars, aes(x = reorder(row.names(mtcars), mpg), y = mpg)) +
  geom_bar(aes(fill = cyl),stat = "identity") + 
  coord_flip() +
  xlab("Car Names") + 
  ylab("Miles per Gallon") + 
  ggtitle("MPG by Car Name")
```
## Plotly 

Plotly's R graphing library makes interactive, publication-quality graphs online. Examples of how to make line plots, scatter plots, area charts, bar charts, error bars, box plots, histograms, heatmaps, subplots, multiple-axes, and 3D (WebGL based) charts. The R package plotly, a high-level interface to the open source JavaScript graphing library plotly.js (which powers plot.ly).


Plotly provides graphics that are:

    Interactive by default.
    Language agnostic.
    Instead of saving as images, charts are serialized as JSON and can 
    be read in R, Python Julia, MATLAB, or JavaScript.
    Embeddable & editable on the Web.
    Vector export for print publication.
    
    
## Traces

A plotly visualization is composed of one (or more) trace(s), and every trace has a type. The default trace type, “scatter”, can be used to draw a large amount of geometries, and actually powers many of the add_*() functions such as add_markers(), add_lines(), add_paths(), add_segments(), add_ribbons(), and add_polygons(). Among other things, these functions make assumptions about the mode of the scatter trace, but any valid attribute(s) listed under the scatter section of the figure reference may be used to override defaults.

The plot_ly() function has a number of arguments that make it easier to scale data values to visual aesthetics (e.g., color/colors, symbol/symbols, linetype/linetypes, size/sizes). These arguments are unique to the R package and dynamically determine what objects in the figure reference to populate (e.g., marker.color vs line.color). 

## Convert ggplot created graphics:
    
Plotly can convert ggplot2 graphs to interactive versions with ggplotly()

With one line of code, convert your ggplot2 graph to an interactive, Web embeddable 
version.

```{r}
library(plotly)
mtcars$am <- factor(mtcars$am,labels=c("A","M"))
p <- ggplot(mtcars, aes(x=wt,y=mpg,color=am)) + 
  geom_point() + 
  ggtitle("MPG vs Wt")

p
```

Let's make this interactive:

```{r}
p <- ggplotly(p)
p
```

```{r}
p <- ggplotly(p)
layout(p,dragmode="pan")
```

## More Common Plots with Plotly

```{r}
mtcars <- mtcars %>% mutate(cyl=factor(cyl))
plot_ly(mtcars, x=~cyl, y=~mpg, type="box")
```


